---
title: "Reproducible Research Project 1"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r, echo = TRUE, message = FALSE, warning= FALSE}
library(dplyr)
library(mice)
library(ggplot2)
```

## Loading and preprocessing the data

In this step, I'll be loading the dataset and performing basic preprocessing on it so that it can be used for further analysis

```{r load, echo = TRUE}
activity <- read.csv("activity.csv")
activity <- transform(activity, date = as.Date(date))
head(activity)
```

## What is mean total number of steps taken per day?

To find the mean number of steps per day, first we group the dataset using `dplyr` package and then summarise the data by finding mean number of steps per day

```{r, echo=TRUE, message = FALSE}
activity_sub <- subset(activity, !is.na(steps))
activity_by_day <- group_by(activity_sub, date)
activity_by_day <- summarise(activity_by_day, total_steps = sum(steps))
head(activity_by_day)
with(activity_by_day, hist(total_steps, col = "#d9a61c", xlab = "Total number of steps by day", main = "Histogram of Total steps by day"))
abline(v = mean(activity_by_day$total_steps), lw = 2, col = "red")
cat("The mean number of steps per day are: ",mean(activity_by_day$total_steps))
cat("The median number of steps per day are: ",median(activity_by_day$total_steps))
```

Since mean and median values are very close, they are represented by the same vertical red line in the graph.

## What is the average daily activity pattern?
First the data has to be prepared in order to do analysis on daily activity pattern. Therefore, the data is grouped by 5 minute interval labels and then it is summarised by finding average number of steps per 5 minute time label, averaged across all the days.
```{r, echo=TRUE, message= FALSE}
activity_by_interval <- group_by(activity_sub, interval)
activity_by_interval <- summarise(activity_by_interval, mean_steps = mean(steps))
head(activity_by_interval)
with(activity_by_interval, plot(interval, mean_steps, type='l', col = "#0080FF", xlab = "5 minute intervals", ylab = "Average number of steps", main = "Average daily activity pattern"))
```
The 5 minute interval having maximum number of steps can be found as:
```{r, echo=TRUE}
activity_by_interval[activity_by_interval$mean_steps == max(activity_by_interval$mean_steps),]
```
Thus, the 5 minute interval labeled by `835` has the maximum number of steps on average, which satisfies the pattern shown in the plot.

## Imputing missing values
```{r, echo = TRUE}
sum(is.na(activity))
```
This shows the number of missing values in the dataset.
```{r, echo = TRUE, message= FALSE}
imp <- mice(activity, where = is.na(activity))
imputed_activity <- complete(imp)
sum(is.na(imputed_activity))
```
Using the mice package to perform imputations. After the process is complete, it can be seen that there are zero missing values.

## Histogram of the total number of steps after imputing data
After imputing the data, we can create another histogram to see things more clearly.
``` {r, echo = TRUE}
by_day <- group_by(imputed_activity, date)
by_day <- summarise(by_day, total_steps = sum(steps))
with(by_day, hist(total_steps, col = "#d9a61c", xlab = "Total number of steps by day", main = "Histogram of Total steps by day"))
```

## Are there differences in activity patterns between weekdays and weekends?
To perform this analysis, first the data has to be prepared.
```{r, echo=TRUE}
imputed_activity <- transform(imputed_activity, day_factor = factor(ifelse(weekdays(date) %in% c("Saturday","Sunday"), "weekend","weekday")))
head(imputed_activity)
```
To calculate average number of steps, average across weekdays or weekends, the data has to be grouped on the basis of two attributes, `interval` and `day_factor`.
```{r, echo=TRUE}
grouped <- group_by(imputed_activity, interval, day_factor)
grouped <- summarise(grouped, avg_steps = mean(steps))
head(grouped)
```
Now the plot can be made.
``` {r, echo=TRUE}
g <- ggplot(grouped, aes(x = interval, y = avg_steps))
g + geom_line(color='steelblue') + facet_grid(day_factor ~ .) + labs(title = "Activity patterns between weekdays and weekends", y = "Number of Steps")
```